---
title: "Traffic Accidents in Munich — Dataset and Research Questions"
format:
  html:
    embed-resources: true
    output-file: "index.html"
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(dplyr)

raw <- read.csv("data/raw/verkehrsunfaelle.csv", encoding = "UTF-8", stringsAsFactors = FALSE)

clean <- raw %>%
  filter(MONAT != "Summe") %>%
  mutate(
    jahr  = JAHR,
    monat = as.integer(substr(MONAT, 5, 6)),
    wert  = WERT
  ) %>%
  filter(!is.na(wert)) %>%
  select(monatszahl = MONATSZAHL, auspraegung = AUSPRAEGUNG, jahr, monat, wert)
```

## 0. Group and Repository

**Group members:** Selma, Yannick

**GitHub repository:** [StatProg2-Group-Project](https://github.com/SReynartz/StatProg2-Group-Project)

## 1. Source and Licence

This dataset was published by the Open Data Portal of the City of Munich (Statistical Bureau of Munich). It is available as *Monatszahlen Verkehrsunfälle* and can be downloaded in CSV format; the portal also provides other formats such as XML and JSON. The data is released under the [Data Licence Germany – Attribution – Version 2.0](http://www.govdata.de/dl-de/by-2-0) (dl-by-de), which allows reuse as long as the source is credited. The file used for this analysis was accessed on 2026-06-07.

## 2. What Is in It

The dataset contains 2450 rows and 10 columns. Each row represents one monthly record for a specific traffic accident category in Munich, uniquely identified by the combination of `MONATSZAHL` (accident category, e.g. *Alkoholunfälle*), `AUSPRAEGUNG` (sub-category, e.g. *insgesamt*), `JAHR` (year), and `MONAT` (year-month code); the `_id` column is a surrogate key for this combination. The main variables used in our analysis are `MONATSZAHL`, `AUSPRAEGUNG`, `JAHR`, `MONAT`, and `WERT` (the monthly count). The remaining columns are pre-computed comparison indicators provided by the data publisher: `VORJAHRESWERT` is the equivalent count for the same month one year earlier; `VERAEND_VORMONAT_PROZENT` and `VERAEND_VORJAHRESMONAT_PROZENT` are percentage changes relative to the previous month and the same month in the prior year, respectively; and `ZWOELF_MONATE_MITTELWERT` is a 12-month rolling average. These derived columns are not used in our analysis.

### Initial data check

```{r}
#| label: ida-glimpse
dplyr::glimpse(raw)
```

```{r}
#| label: ida-missing
#| tbl-cap: "Missing values per column in the raw dataset"
knitr::kable(t(colSums(is.na(raw))))
```

```{r}
#| label: fig-ida-yearly-trend
#| fig-cap: "Yearly totals by accident category (raw, unfiltered data)"

yearly <- raw %>%
  filter(MONAT == "Summe", AUSPRAEGUNG == "insgesamt", !is.na(WERT))

plot(NULL, xlim = range(yearly$JAHR), ylim = range(yearly$WERT),
     xlab = "Year", ylab = "Total accidents per year")
kategorien <- unique(yearly$MONATSZAHL)
farben <- c("steelblue", "tomato", "darkgreen")
for (i in seq_along(kategorien)) {
  d <- yearly[yearly$MONATSZAHL == kategorien[i], ]
  d <- d[order(d$JAHR), ]
  lines(d$JAHR, d$WERT, type = "b", pch = 16, col = farben[i], lwd = 2)
}
legend("topright", legend = kategorien, col = farben, lwd = 2, pch = 16, bty = "n")
```

**Data quality notes:**

- **Missing values:** `WERT` has missing values, especially for the most recent year (2026), where monthly figures have not yet been recorded.
- **Inconsistent formatting:** the `MONAT` column mixes the `YYYYMM` format (e.g. `"202601"`) with the string `"Summe"` for annual totals, which violates the tidy-data principle of one observation per row.
- **Mixed granularity:** because monthly counts and annual totals are stored in the same `WERT` column, naive aggregation without filtering out `"Summe"` rows would silently double-count accidents. All analyses below therefore explicitly drop `MONAT == "Summe"` rows (see `scripts/01_clean_data.R`) before computing monthly statistics.

## 3. Research Questions

### Exploratory

**How do alcohol-related accidents vary over the course of the year? Are there seasonal patterns, and how does their share of all traffic accidents change across months?**

This is mainly an exploratory question because it looks for recurring patterns and possible seasonal structure in alcohol-related accidents. It also has a descriptive component because it examines the monthly share of alcohol-related accidents relative to total traffic accidents.

**Output (variables used:** `MONATSZAHL`, `AUSPRAEGUNG`, `JAHR`, `MONAT`, `WERT`**):**

```{r}
#| label: fig-alcohol-seasonality
#| fig-cap: "Figure 1: Seasonal pattern of alcohol-related accidents in Munich (mean monthly count, averaged across all years)"

alkohol <- clean %>%
  filter(monatszahl == "Alkoholunfälle", auspraegung == "insgesamt") %>%
  select(jahr, monat, alkohol_wert = wert)

gesamt <- clean %>%
  filter(monatszahl == "Verkehrsunfälle", auspraegung == "insgesamt") %>%
  select(jahr, monat, gesamt_wert = wert)

merged <- inner_join(alkohol, gesamt, by = c("jahr", "monat")) %>%
  mutate(share_pct = 100 * alkohol_wert / gesamt_wert)

fig1_data <- merged %>%
  group_by(monat) %>%
  summarise(mean_alkohol = mean(alkohol_wert), .groups = "drop")

plot(
  fig1_data$monat, fig1_data$mean_alkohol,
  type = "b", pch = 16, col = "steelblue", lwd = 2,
  xaxt = "n", xlab = "Month", ylab = "Mean number of alcohol-related accidents"
)
axis(1, at = 1:12)
```

Figure 1 shows a clear seasonal pattern: alcohol-related accidents are at their lowest in winter (January–February, around 31–32 per month on average) and rise steadily through spring and summer, peaking in July (\~55) and again in September (\~57). This suggests that warmer months with more outdoor activity and nightlife are associated with higher alcohol-related accident rates, while the dip in August relative to the surrounding months may reflect holiday absences reducing commuter traffic.

```{r}
#| label: tbl-alcohol-share
#| tbl-cap: "Table 1: Monthly average share (%) of alcohol-related accidents relative to all traffic accidents"

table1 <- merged %>%
  group_by(monat) %>%
  summarise(
    mean_alkohol_unfaelle = round(mean(alkohol_wert), 1),
    mean_gesamt_unfaelle  = round(mean(gesamt_wert), 1),
    mean_share_pct        = round(mean(share_pct), 2),
    .groups = "drop"
  )

knitr::kable(table1)
```

Table 1 confirms that the seasonal pattern in Figure 1 is not simply driven by higher overall accident volumes in summer: the *share* of alcohol-related accidents also peaks in late summer and early autumn (August 1.51 %, September 1.54 %), while winter months (February–March, around 1.0 %) have a proportionally lower alcohol share. This means alcohol-related accidents are overrepresented in summer relative to total accidents, not merely more frequent in absolute terms.

### Predictive

**Can the number of traffic accidents in the following month be predicted using a simple regression model that accounts for long-term trend and seasonality?**

This model is used for **short-term forecasting**, not causal explanation: the goal is to evaluate whether a parsimonious model can anticipate next-month accident volumes well enough to be useful for planning purposes. The coefficients for the month dummies quantify average seasonal effects relative to January, and the linear trend term captures a gradual long-run change in accident counts, but neither should be interpreted as causal mechanisms.

**Output (variables used:** `MONATSZAHL`, `AUSPRAEGUNG`, `JAHR`, `MONAT`, `WERT`**):**

```{r}
#| label: tbl-regression
#| tbl-cap: "Table 2: Linear regression of total monthly accident counts on a linear time trend and calendar-month dummies"

gesamt_ts <- clean %>%
  filter(monatszahl == "Verkehrsunfälle", auspraegung == "insgesamt") %>%
  arrange(jahr, monat) %>%
  mutate(t = row_number(), monat_f = factor(monat, levels = 1:12))

n_test <- 12
n <- nrow(gesamt_ts)
train <- gesamt_ts[1:(n - n_test), ]
test  <- gesamt_ts[(n - n_test + 1):n, ]

model <- lm(wert ~ t + monat_f, data = train)

knitr::kable(broom::tidy(model), digits = 3)
```

Table 2 shows the fitted regression coefficients. The intercept (\~2932) represents the model's baseline for January of the first year in the training period. The trend coefficient (\~0.53 per month, approximately 6 additional accidents per year) indicates a modest long-run upward drift. The month dummies reveal strong seasonality: July carries the largest positive effect (+1043 accidents relative to January), followed by October (+940) and September (+728), consistent with the exploratory findings. February and August dummies are small and statistically weak, suggesting those months are close to the January baseline once trend is accounted for.

```{r}
#| label: fig-predicted-vs-actual
#| fig-cap: "Figure 2: Predicted vs. actual monthly accident counts on the held-out test period (last 12 months)"

pred <- predict(model, newdata = test)
rmse <- sqrt(mean((test$wert - pred)^2))

plot(
  test$t, test$wert,
  type = "b", pch = 16, col = "black", lwd = 2,
  xlab = "Time index (test period, last 12 months)", ylab = "Monthly accident count",
  sub = sprintf("Test-set RMSE = %.1f", rmse),
  ylim = range(c(test$wert, pred))
)
lines(test$t, pred, type = "b", pch = 17, col = "tomato", lwd = 2)
legend("topleft", legend = c("Actual", "Predicted"), col = c("black", "tomato"), pch = c(16, 17), lwd = 2, bty = "n")
```

Figure 2 shows that the model tracks the seasonal shape of the test period reasonably well, capturing the winter trough and summer peaks. The test-set RMSE of approximately 233 represents a relative error of around 6–7 % against typical monthly totals of 3000–4500 accidents, which is acceptable for a simple two-parameter model (trend + month dummies). The model tends to underestimate the highest peaks (notably the July spike), suggesting that a more flexible approach — such as adding interaction terms or using a non-linear trend — could improve accuracy at the extremes.

## 4. Analysis Plan

Each step is trackable as a separate GitHub Issue and assignable to one group member. Steps 1–5 (the core analysis behind Section 3) are already complete; the remaining weeks focus on presentation and polishing.

1.  ✅ **Data cleaning** (Yannick) — load the raw CSV, filter to the relevant `MONATSZAHL`/`AUSPRAEGUNG` categories ("Alkoholunfälle" and "Verkehrsunfälle" / "insgesamt"), drop non-monthly rows (e.g. `MONAT = "Summe"`), and export a tidy long-format data frame with `JAHR`, `MONAT`, `MONATSZAHL`, `WERT`.
2.  ✅ **Initial data exploration and quality checks** (Yannick) — `glimpse()`/`summary()`, missing-value counts, and documentation of data-quality issues (see Section 2).
3.  ✅ **Exploratory analysis: seasonality of alcohol-related accidents** (Selma) — compute monthly means and the monthly % share of alcohol-related accidents relative to total accidents; produce *Figure 1* and *Table 1*.
4.  ✅ **Feature engineering and model fitting for the predictive question** (Selma) — build a linear time-trend variable and calendar-month dummies from `JAHR`/`MONAT`, split into training/held-out test period (last 12 months), fit the regression and produce *Table 2*.
5.  ✅ **Predictive evaluation** (Selma) — generate test-period predictions and produce *Figure 2* (predicted vs. actual, with RMSE).
6.  ✅ **Visualisation polish** (Selma & Yannick) — ~~restyle Figures 1–2 (e.g. consistent `ggplot2` theme per Section 6 workflow conventions), refine table formatting,~~ write up interpretation text for each figure/table.
7.  ✅ **Final presentation assembly** (Selma & Yannick) — integrate the polished figures/tables into the final report/slide format for hand-in.

## 5. Context and Target Audience

The project is situated in a public policy and urban planning context. The target audience includes local policymakers, city administrators, urban mobility planners, and road safety stakeholders in Munich. For this audience, suitable presentation formats include concise reports, policy briefs, slide presentations, dashboards, and clear visualizations such as time series plots, seasonal comparisons, and simple forecasts.

## 6. Workflow Organisation

- **Canonical cleaning script:** `scripts/01_clean_data.R` (reads `data/raw/verkehrsunfaelle.csv`, writes `data/clean_data.csv`). Run from the project root.
- **Code style:** tidyverse style, enforced with `styler`.
- **Packages:** `renv` for version reproducibility, `styler`, `knitr`, `readr`, `ggplot2` (exploratory/predictive chunks in this document currently use base R plotting for local-rendering robustness; restyling to `ggplot2` is tracked in Analysis Plan step 6).
- **Git workflow:** each group member works on their own branch; regular PRs/merges keep members up to date. Tasks are split among members but not necessarily placed in new branches.
- **`.gitignore`:** excludes `.Rproj.user`, `.Rhistory`, `.RData`, `.Ruserdata`, `.env` (if applicable), and rendered outputs.
