---
title: "project-proposal"
format: html
editor: visual
---

# Dataset & research questions

## 1. Source and Licence

This dataset was published by the Open Data Portal of the City of Munich. It is available as *Monatszahlen Verkehrsunfälle* and can be downloaded in CSV format; the portal also provides other formats such as XML and JSON. The data is released under the Datenlizenz Deutschland Namensnennung 2.0 (dl-by-de), which allows reuse as long as the source is credited.

## 2. What Is in It

The dataset contains 2450 rows and 10 columns. Each row represents one monthly record for a specific traffic accident category in Munich. The table includes the year and month of the observation, the recorded value, and additional indicators such as previous-year values and percentage changes over time. The main variables are `_id`, `MONATSZAHL`, `AUSPRAEGUNG`, `JAHR`, `MONAT`, `WERT`, `VORJAHRESWERT`, `VERAEND_VORMONAT_PROZENT`, `VERAEND_VORJAHRESMONAT_PROZENT`, and `ZWOELF_MONATE_MITTELWERT`.

## 3. Research Questions

### Descriptive

**How have monthly traffic accident counts in Munich developed over time?**

This is a descriptive question because it focuses on summarizing and visualizing long-term developments in the monthly accident data.

### Exploratory

**How do alcohol-related accidents vary over the course of the year? Are there seasonal patterns, and how does their share of all traffic accidents change across months?**

This is mainly an exploratory question because it looks for recurring patterns and possible seasonal structure in alcohol-related accidents. It also has a descriptive component because it examines the monthly share of alcohol-related accidents relative to total traffic accidents.

### Predictive

**Can the number of traffic accidents in the following month be predicted using a simple regression model that accounts for long-term trend and seasonality?**

This question aims to evaluate whether a basic predictive model can capture the temporal structure of the data well enough to generate short-term forecasts.

## 4. Context and Target Audience

The project is situated in a public policy and urban planning context. The target audience includes local policymakers, city administrators, urban mobility planners, and road safety stakeholders in Munich. For this audience, suitable presentation formats include concise reports, policy briefs, slide presentations, dashboards, and clear visualizations such as time series plots, seasonal comparisons, and simple forecasts.

# Initial data analysis:

### 1. `glimpse()`

```{r}
library(dplyr)
library(readr)
traffic_accidents <- read_csv("data/verkehrsunfaelle.csv")
glimpse(traffic_accidents)
```

### 2. Summary statistics for key variables

```{r}
# Summary statistics
summary(traffic_accidents)
```

```{r}
# Count missing values per column
colSums(is.na(traffic_accidents))
```

### 3. Exploratory Plots

```{r}
library(ggplot2)

# Plot 1: Distribution of monthly accident values by category
traffic_accidents |>  
  filter(monat != "Summe", !is.na(wert)) |> 
  ggplot(aes(x = wert, fill = monatszahl)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~ monatszahl, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Monthly Accidents by Category",
       x = "Number of Accidents (wert)",
       y = "Count")
```

```{r}
# Plot 2: Yearly trend of total accidents by category
traffic_accidents |>  
  filter(monat == "Summe", auspraegung == "insgesamt", !is.na(wert)) |> 
  ggplot(aes(x = jahr, y = wert, color = monatszahl)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Yearly Trend of Total Accidents by Category",
       x = "Year",
       y = "Total Number of Accidents (wert)")
```

### 4. Notes on data quality issues

- **Missing values**: `wert` has 84 NAs (especially for the year 2026 where values seem to be not yet recorded). Other calculated columns like `veraend_vormonat_prozent` also have NAs (up to 274).

- **Inconsistent formatting**: The `monat` column mixes YYYYMM format (e.g., "202601") with the string "Summe" to represent annual totals, violating the tidy data principle.

- **Unexpected values**: Because monthly counts and annual totals ("Summe") are kept in the same column (`wert`), the range of `wert` is extremely skewed (0 to 46,988), which could bias naive aggregations or plots. The year is duplicated in both the `jahr` and `monat` columns.

# **Analysis Plan**

- Variables**:** Number of accidents (*WERT*, numerical), year (*JAHR*, numerical), month (*MONAT*, numerical), accident type (*MONATSZAHL*, categorical), and accident severity/category (*AUSPRAEGUNG*, categorical).

- Method**:** Time-series and seasonal analysis to examine long-term trends, monthly patterns. A simple regression model with trend and seasonal variables can be used to forecast accident counts for the following month

- Data Cleaning**:** Convert month and year into a proper date variable, check for missing values, and aggregate data by year, month, or accident category where needed.

- Risks/Uncertainties**:** Missing or incomplete records, short observation periods for some categories, seasonal fluctuations, and external influences that are not included in the dataset.

# Workflow organisation:

- Code Style:

  - tidyverse style

  - enforced with styler

- Packages:

  - renv for version reproducibility

  - styler

  - knitr

  - readr

  - ggplot2

- Git workflow:

  - each group member has his own branch

  - regular pr/merges ensure members stay up to date

  - the tasks will be split among the members, but not necessarly be placed in new branches

- .gitignore:

  - if applicable `.env` will be excluded

  - raw data files

  - rendered outputs
