---
title: "project-proposal"
format: html
editor: visual
---

## Initial data analysis:

### 1. `glimpse()`

```{r}
library(dplyr)
library(readr)
traffic_accidents <- read_csv("data/verkehrsunfaelle.csv")
glimpse(traffic_accidents)
```

### 2. Summary statistics for key variables

```{r}
# Summary statistics
summary(traffic_accidents)
```

```{r}
# Count missing values per column
colSums(is.na(traffic_accidents))
```

### 3. Exploratory Plots

```{r}
library(ggplot2)

# Plot 1: Distribution of monthly accident values by category
traffic_accidents |>  
  filter(monat != "Summe", !is.na(wert)) |> 
  ggplot(aes(x = wert, fill = monatszahl)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~ monatszahl, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Monthly Accidents by Category",
       x = "Number of Accidents (wert)",
       y = "Count")
```

```{r}
# Plot 2: Yearly trend of total accidents by category
traffic_accidents |>  
  filter(monat == "Summe", auspraegung == "insgesamt", !is.na(wert)) |> 
  ggplot(aes(x = jahr, y = wert, color = monatszahl)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Yearly Trend of Total Accidents by Category",
       x = "Year",
       y = "Total Number of Accidents (wert)")
```

### 4. Notes on data quality issues

- **Missing values**: `wert` has 84 NAs (especially for the year 2026 where values seem to be not yet recorded). Other calculated columns like `veraend_vormonat_prozent` also have NAs (up to 274).
- **Inconsistent formatting**: The `monat` column mixes YYYYMM format (e.g., "202601") with the string "Summe" to represent annual totals, violating the tidy data principle.
- **Unexpected values**: Because monthly counts and annual totals ("Summe") are kept in the same column (`wert`), the range of `wert` is extremely skewed (0 to 46,988), which could bias naive aggregations or plots. The year is duplicated in both the `jahr` and `monat` columns.

## Workflow organisation:

- Code Style:

  - tidyverse style

  - enforced with styler

- Packages:

  - renv for version reproducibility

  - styler

  - knitr

- Git workflow:

  - each group member has his own branch

  - regular pr/merges ensure members stay up to date

  - the tasks will be split among the members, but not necessarly be placed in new branches

- .gitignore:

  - if applicable `.env` will be excluded

  - raw data files

  - rendered outputs
